Constructing a universe for the setoid model

### 2021-01-22 paper.tex
spacing

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
{\small\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | \sigma | &: | \GG | \to | \GD | \\
    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\}\vspace{-1em}

@@

As an example, consider the following definition of a small universe with
bool and $\Pi$.

{\small\vspace{-0.4em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}}\vspace{-1em}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
{\small\vspace{-0.4em}
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}}\vspace{-2em}

@@

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible types in
$\mType$. Hence this kind of construction requires a metatheory with at least
one universe.

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations: \\

{\small\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{data} \ \setoidU &: \mType_1 \\
  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}}\vspace{0.4em}

@@

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
{\vspace{-0.4em}\small\begin{align*}
  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}}\vspace{-1.6em}

We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined: $\textsf{refl}_\El$ alone,
$\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
$\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% The other functions are then recursively defined:
% \begin{itemize}
% \item $\textsf{refl}_\El$ alone,
% \item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
% \item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% \end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$:

{\small\vspace{-1em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}}\vspace{1em}

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
{\small\vspace{-0.6em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  \textsf{refl}\ \Uty &\defeq \reflu \\
  &\ldots \\
  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
  &\ldots \\
  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}}
\\

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
{\small\vspace{-0.2em}
\begin{minipage}{0.4\textwidth}
\begin{align*}
  \bullet_0 &: \textsf{Con}_0 \\
  \blank,_0\blank &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  \iota_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \\
  \Pi_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  \bullet_1 &: \textsf{Con}_1\ \bullet_0 \\
  \blank,_1\blank &: \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  \iota_1 &: \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  \Pi_1 &: \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}}\vspace{0.4em}

>>>>

Substitutions are interpreted as functors between the corresponding setoids,
whereas terms of type $A$ in context $\GG$ are sections of the type seen as a
setoid fibration $\GG . A \to \GG$. Note that we only need to include components
for the functorial action on objects and morphisms, since the functor laws
follow from proof-irrelevance in the metatheory, and thus hold definitionally.
\\
{\small\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | \sigma | &: | \GG | \to | \GD | \\
    \sigma^\sim &: \GG^\sim\, \rho_0\, \rho_1 \to \GD^\sim\, (|\sigma|\rho_0)\, (|\sigma|\rho_1)
  \end{aligned}
}{
  \sigma : \Sub\ \GG\ \GD
}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\[
\infer{
  \begin{aligned}
    | t | &: (\gamma : | \GG |) \to | A | \, \gamma \\
    t^\sim &: (p : \GG^\sim\ \gamma_0\ \gamma_1) \to A^\sim\ p\ (|t|\gamma_0)\ (|t|\gamma_1)
  \end{aligned}
}{
  t : \Tm\ \GG\ A
}
\]
\end{minipage}
\\}

@@

An example is the following definition of a small universe with
bool and $\Pi$.

{\small\vspace{-0.5em}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \textsf{data}\ \U\ : \mType \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
\end{align*}
\end{minipage}
\begin{minipage}{0.45\textwidth}
\begin{align*}
  & \El\ : \U \to \mType \\
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) \defeq (a : \El\ A) \to \El\ (B\ a) \\
\end{align*}
\end{minipage}}\vspace{-1em}

@@

In our case, we can encode $\U$ as an inductive type $\inU$ that
\emph{carves out} all types in $\mType$ that are in the image of $\El$. In other
words, $\inU$ is a predicate that holds for any type that would have been
obtained via $\El$ in the inductive-recursive definition. As $\El$ is indexed by
the type of codes, the definition of $\inU$ quite expectedly reflects the
inductive structure of codes.
{\small
\begin{alignat*}{10}
  & \textsf{data}\ \inU && : \mType \to \mType_1 \\
  & \quad \textsf{inBool} && : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi} && :  \inU\ A \to ((a : A) \to \inU\ (B\ a)) \to \inU\ ((a : A) \to (B\ a))
\end{alignat*}}\vspace{-1em}

@@

Note that this construction gives rise to a universe in $\mType_1$, rather than
$\mType$, since the definition of $\U$ quantifies over all possible types in
$\mType$. Hence this kind of construction requires a metatheory with at least
one universe.
\vspace{-0.3em}

@@

To construct the universe of setoids in the setoid model, we first of all need
to define a type $\Uty : \Ty\ \GG$ for every $\GG : \Con$, and for every $A :
\Tm\ \GG\ \Uty$ a type $\mathds{E}l\ A : \Ty\ \GG$. Recalling
Section~\ref{sec:setoid-model}, these are essentially record types made of several
components. Since $\Uty$ is a closed type, it requires the same data of a
setoid; in particular, we need a type of codes together with an equivalence
relation reflecting equality of codes, in addition to proofs that these are
indeed equivalence relations:\vspace{-1em} \\
{\small\begin{minipage}{0.4\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{data} \ \setoidU &: \mType_1 \\
  \blank\sim_\setoidU\blank &: \setoidU \to \setoidU \to \mProp_1
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-0.2em}
\begin{align*}
  \textsf{refl}_\setoidU &: (A : \setoidU) \to \equ{A}{A} \\
  \textsf{sym}_\setoidU &: \equ{A}{B} \to \equ{B}{A} \\
  \textsf{trans}_\setoidU &: \equ{A}{B} \to \equ{B}{C} \to \equ{A}{C}
\end{align*}
\end{minipage}}\vspace{0.4em}

@@

Note that $\eqel{\blank}{\blank}{\blank}$ is indexed over equality on the universe, because
$\El$ is a displayed setoid over $\setoidU$, hence in particular it must respect the setoid
equality of $\setoidU$.
%
We also require data and proofs that make sure we get setoids out of $\El$:
{\small\begin{align*}
  \textsf{refl}_\El &: (A : \setoidU) (x : \El\ A) \to \eqel{\reflu\ A}{x}{x} \\
  \textsf{sym}_\El &: \eqel{p}{x}{x'} \to \eqel{\symu\ p}{x'}{x} \\
  \textsf{trans}_\El &: \eqel{p}{x}{x'} \to \eqel{q}{x'}{x''} \to \eqel{\transu\ p\ q}{x}{x''} \\
  \coeel &: \equ{A}{B} \to \El\ A \to \El\ B \\
  \cohel &: (p : \equ{A}{A'}) \ (x : \El\ A) \to \eqel{p}{x}{\coeel\ p\ x}
\end{align*}}\vspace{-1em}

We give an inductive definition of $\setoidU$, mutually with a recursive definition
of the 4 functions
$\blank\sim_\setoidU\blank$, $\textsf{refl}_\setoidU$, $\El$ and $\blank\vdash\blank\sim_\El\blank$.
The other functions are then recursively defined: $\textsf{refl}_\El$ alone,
$\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
$\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% The other functions are then recursively defined:
% \begin{itemize}
% \item $\textsf{refl}_\El$ alone,
% \item $\textsf{sym}_\setoidU$ and $\textsf{sym}_\El$ mutually,
% \item $\textsf{trans}_\setoidU$, $\textsf{trans}_\El$, $\coeel$ and $\cohel$ mutually.
% \end{itemize}
The whole construction is quite long, below we only show the more interesting
definitions of $\setoidU$ and $\El$: \\
{\small\vspace{0em}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \textsf{data}\ \setoidU : \mType_1 \\
  & \quad \textsf{bool} : \setoidU \\
  & \quad \textsf{pi}
  :  (A : \setoidU) (B : \El\ A \to \setoidU)\\
  & \ \qquad \to (\{x\ x' : \El\ A\} \to \textsf{refl}_\setoidU\ A \vdash x \sim_\El x'\\
  & \qquad\quad \to B\ x \sim_\setoidU B\ x')
  \to \setoidU
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  & \El\ \textsf{bool} \defeq \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B\ h) \defeq \\
  & \qquad \Sigma\ (f : (a : \El\ A) \to \El\ (B\ a)) \\
  & \qquad \quad
  (\forall\{x\ x'\}(p : \eqel{\reflu\ A}{x}{x'}) \\
  & \qquad \quad\to \eqel{h\ p}{f\ x}{f\ x'})
\end{align*}
\end{minipage}}\vspace{1em}

@@

We can now give a full definition of the setoid universe, and of $\EL\ A$ for
any $A : \Tm\ \GG\ \Uty$: \\
{\small
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \Uty | &\defeq \lambda\, \gamma . \, \setoidU \\
  \Uty^\sim &\defeq \lambda \,p\, x\, y . \,x \sim_{\setoidU} y \\
  \textsf{refl}\ \Uty &\defeq \reflu \\
  &\ldots \\
  \textsf{coe}\ \Uty &\defeq \lambda\,p\,a . \, a \\
  \textsf{coh}\ \Uty &\defeq \lambda\,p . \, \reflu
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{align*}
  | \EL\ A | &\defeq \lambda \, \gamma . \, \El\ (| A | \,\gamma) \\
  (\EL\ A)^\sim &\defeq \lambda \, p \, x \, y . \, \eqel{A^\sim\,p}{x}{y} \\
  \textsf{refl}\ (\EL\ A) &\defeq \reflel \\
  &\ldots \\
  \textsf{coe}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coe}_{\El}\,(A^\sim\,p) \\
  \textsf{coh}\ (\EL\ A) &\defeq \lambda\,p.\, \textsf{coh}_{\El}\,(A^\sim\,p)
\end{align*}
\end{minipage}}
\\

@@

The definition of the codes and predicate types follows that of the original
inductive-inductive type, and can be derived systematically from it. More
importantly, they can be defined without induction-induction, since although
$\textsf{Con}_0$ and $\textsf{Ty}_0$ are defined mutually, their sorts are not
indexed. \\
{\small
\begin{minipage}{0.4\textwidth}
\begin{align*}
  \bullet_0 &: \textsf{Con}_0 \\
  \blank,_0\blank &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Con}_0 \\
  \iota_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \\
  \Pi_0 &: \textsf{Con}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \to \textsf{Ty}_0 \hspace{2.8em} \\
\end{align*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\vspace{-1.3em}
\begin{alignat*}{10}
  \bullet_1 &: \textsf{Con}_1\ \bullet_0 \\
  \blank,_1\blank &: \forall\{\GG_0\ A_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ A_0 \\
  & \to \textsf{Con}_1\ (\GG_0 \,,_0 A_0) \\
  \iota_1 &: \forall\{\GG_0\} \to \textsf{Con}_1\ \GG_0 \to \textsf{Ty}_1\ \GG_0\ (\iota_0\ \GG_0) \\
  \Pi_1 &: \forall\{\GG_0\ A_0\ B_0\} \to \textsf{Con}_1\ \GG_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ A_0 \to \textsf{Ty}_1\ (\GG_0\, ,_0 A_0)\ B_0 \\
  & \to \textsf{Ty}_1\ \GG_0\ (\Pi_0\ \GG_0\ A_0\ B_0)
\end{alignat*}
\end{minipage}}\vspace{0.4em}
